---
title: "BinFengHW3"
author: "Bin Feng"
output: pdf_document
---
```{r setup}
#include library
require("knitr")
#set working directory
opts_knit$set(root.dir = "~/Stat480/Homework")
```

```{r}
# The following initial code is given by Darren Glosemeyer. The contents of this 
# include segments of code from Chapter 3 up through section 3.5.4. The following 
# lines from previous sections are needed to define sampleSplit, getBoundary,
# and dropAttach.

spamPath = "~/Stat480/RDataScience/SpamAssassinMessages"

dirNames = list.files(path = paste(spamPath, "messages", 
                                   sep = .Platform$file.sep))
fullDirNames = paste(spamPath, "messages", dirNames, 
                     sep = .Platform$file.sep)

indx = c(1:5, 15, 27, 68, 69, 329, 404, 427, 516, 852, 971)
fn = list.files(fullDirNames[1], full.names = TRUE)[indx]
sampleEmail = sapply(fn, readLines)

splitMessage = function(msg) {
  splitPoint = match("", msg)
  header = msg[1:(splitPoint-1)]
  body = msg[ -(1:splitPoint) ]
  return(list(header = header, body = body))
}

sampleSplit = lapply(sampleEmail, splitMessage)

getBoundary = function(header) {
  boundaryIdx = grep("boundary=", header)
  boundary = gsub('"', "", header[boundaryIdx])
  gsub(".*boundary= *([^;]*);?.*", "\\1", boundary)
}

dropAttach = function(body, boundary){
  
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  
  # if there are fewer than 2 beginning boundary strings, 
  # there is on attachment to drop
  if (length(bStringLocs) <= 1) return(body)
  
  # do ending string processing
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  
  # if no ending boundary string, grab contents between the first 
  # two beginning boundary strings as the message body
  if (length(eStringLoc) == 0) 
    return(body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1)])
  
  # typical case of well-formed email with attachments
  # grab contents between first two beginning boundary strings and 
  # add lines after ending boundary string
  n = length(body)
  if (eStringLoc < n) 
    return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), 
                     ( (eStringLoc + 1) : n )) ] )
  
  # fall through case
  # note that the result is the same as the 
  # length(eStringLoc) == 0 case, so code could be simplified by 
  # dropping that case and modifying the eStringLoc < n check to 
  # be 0 < eStringLoc < n
  return( body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1) ])
}

# Set the working directory. This is where R will look for files and save files if a full path is not specified.
# setwd("~/Stat480/RDataScience/Chapter3")
```
# Question 1
Write a function to extract the subject from messages. This is the value for the Subject key (Be sure to allow for capital or lower case s when searching for Subject in the header). The function should take the header from a message and return the subject value, and also handle the case where no Subject key is present in the header.
Demonstrate your function on the headers from the 15 sample emails used in the chapter.
```{r}
headerList = lapply(sampleSplit, function(msg) msg$header)
getSubject = function(header){
  # tolower() to allow both capital and lower cases
  subject.index = grep("subject:", tolower(header))
  # allow no subject line cases
  if (subject.index == 0) return(0)
  # main subject extract part
  subject = header[subject.index]
  sub('Subject: (.*)', "\\1", subject)
}
# test with all 15 sample cases
unname(sapply(headerList, getSubject))

# get the wholeline
# getfull = function(header){
#   subject.index = grep("subject:", tolower(header))
#   subject = header[subject.index]
# }
```

# Question 2
Write a function to count the letters from a message subject line. The function should take in a
message header and return an integer (the number of letters in the subject line). Make use of the function created in exercise 1 and use modular programming to create your function. Again make sure to handle the missing Subject key case.
Demonstrate your function on the 15 sample emails from the chapter.
```{r}
countLetter = function(header){
  subject = unname(sapply(header, getSubject))
  subject.letter = gsub("[^a-zA-Z]+", "", subject) #do we need to get rid of [], ..
  nchar(subject.letter)
}

countLetter(headerList)
```

# Question 3
Write a function to count the number of digits (0, 1, 2, 3, 4, 5, 6, 7, 8, or 9) in an email message body. The function should take a split email message (the result from the splitMessage function in the chapter) and return the number of digits in the body. Drop attachments and only count digits not in attachments.
Again, show the results of your function for the sample emails.
```{r}

countNumber = function(sample){
  headerList = lapply(sample, function(msg) msg$header)
  bodyList = lapply(sample, function(msg) msg$body)
  
  #from course code 
  # determine which messages have attachments
  hasAttach = sapply(headerList, function(header) {
    CTloc = grep("Content-Type", header)
    if (length(CTloc) == 0) return(0)
    multi = grep("multi", tolower(header[CTloc])) 
    if (length(multi) == 0) return(0)
    multi
  })
  
  hasAttach = which(hasAttach > 0)
  
  # find boundary strings for messages with attachments
  boundaries = sapply(headerList[hasAttach], getBoundary)
  
  bodyList[hasAttach] = mapply(dropAttach, bodyList[hasAttach], 
                               boundaries, SIMPLIFY = FALSE)
  
  subject.number = gsub("[^0-9]+", "", bodyList) #do we need to get rid of [], ..
  nchar(subject.number)
}

countNumber(sampleSplit)
```

# Question 4
Write a function to obtain the number of digits in each email in a directory (e.g. easy_ham, easy_ham_2, etc.) that relies on the function written in exercise 3. The function should take a directory name as its input and return a vector of the digits counts for messages in that directory as its output. (Note: the initial processing from processAllWords in section 3.5.4 of the text is a good example for processing a directory of messages.)
Obtain basic summary statistics for the digits per message for each of the 5 email directories, and comment on differences in digits per message in the 5 directories. What do these differences suggest about using number of digits in a message as a basis for classifying spam and ham messages?
```{r}
countAll = function(dirName){
  # read all files in the directory
  fileNames = list.files(dirName, full.names = TRUE)
  # drop files that are not email, i.e., cmds
  notEmail = grep("cmds$", fileNames)
  if ( length(notEmail) > 0) fileNames = fileNames[ - notEmail ]
  
  messages = lapply(fileNames, readLines, encoding = "latin1")
  
  # split header and body
  emailSplit = lapply(messages, splitMessage)
  
  countNumber(emailSplit)
}

countAll(fullDirNames[1]) #incomplete final line found on...
```
