---
title: "STAT480_Homework_4"
author: "Bin Feng"
output: pdf_document
---

```{r setup}
#include library
require("knitr")
#set working directory
opts_knit$set(root.dir = "~/Stat480/Homework")
```
# Setup
```{r}
#run setup code from another .R file
source("HW4Setup.R")
```
# Question 1
Create a function computeMsgLLR2 which implements the log of ratios of products of probabilities formula for the log likelihood ratio statistic. Compare the results from this definition with the results from the computeMsgLLR function used in the text which used the sum of differences of log probabilities. 
Specifically, compare accuracy for this formula compared to the one used in class (Hint: to estimate relative accuracy you should look at (observed-expected)/expected, and treat the results from computeMsgLLR2 as observed and the results from computeMsgLLR as expected) and note any issues that arise with non-representable numbers (e.g. very large or very small intermediate results that result in infinite, incorrect 0, or not a number results from your function).
```{r}
# from course code
computeMsgLLR = function(words, freqTable) 
{
  # Discards words not in training data.
  words = words[!is.na(match(words, colnames(freqTable)))]
  
  # Find which words are present
  present = colnames(freqTable) %in% words
  
  sum(freqTable["presentLogOdds", present]) +
    sum(freqTable["absentLogOdds", !present])
}

# not + .5 because division evaluated first
computeFreqs2 =
  function(wordsList, spam, bow = unique(unlist(wordsList)))
  {
    # create a matrix for spam, ham, and log odds
    wordTable = matrix(0.5, nrow = 4, ncol = length(bow), 
                       dimnames = list(c("spam", "ham", 
                                         "presentLogOdds", 
                                         "absentLogOdds"),  bow))
    
    # For each spam message, add 1/2 to counts for words in message
    counts.spam = table(unlist(lapply(wordsList[spam], unique)))
    wordTable["spam", names(counts.spam)] = counts.spam
    
    # Similarly for ham messages
    counts.ham = table(unlist(lapply(wordsList[!spam], unique)))  
    wordTable["ham", names(counts.ham)] = counts.ham 
    
    
    # Find the total number of spam and ham
    numSpam = sum(spam)
    numHam = length(spam) - numSpam
    
    # Prob(word|spam) and Prob(word | ham)
    wordTable["spam", ] = wordTable["spam", ]/(numSpam)
    wordTable["ham", ] = wordTable["ham", ]/(numHam)
    
    # log odds
    wordTable["presentLogOdds", ] = 
      wordTable["spam",] / wordTable["ham", ]
    wordTable["absentLogOdds", ] = 
      (1 - wordTable["spam", ]) / (1 -wordTable["ham", ])
    
    invisible(wordTable)
  }

computeMsgLLR2 = function(words, freqTable) 
{
  # Discards words not in training data.
  words = words[!is.na(match(words, colnames(freqTable)))]
  
  # Find which words are present
  present = colnames(freqTable) %in% words
  
  log(prod(freqTable["presentLogOdds", present])) +
    log(prod(freqTable["absentLogOdds", !present]))
}

trainTable2 = computeFreqs2(trainMsgWords, trainIsSpam) 

testLLR1 = sapply(testMsgWords, computeMsgLLR, trainTable)
testLLR2 = sapply(testMsgWords, computeMsgLLR2, trainTable2) #how does this sapply works?

relative.accuracy = (testLLR2 - testLLR1) / testLLR1
```

# Question 2
Do exercise Q.13 from page 167 of Data Science in R: A Case Studies Approach to Computational Reasoning and Problem Solving, by Deborah Nolan and Duncan Temple Lang. Within the exercise, construct two functions: one that counts the number of yelling lines, and one that gives the percentage.
```{r}
#body or entire text??
isYelling.num = function(msg) {
  body = msg$body
  if(length(body) == 0) return("no body")
  else{
    # remove non-alpha characters
    body = gsub("[^[:alpha:]]", "", body)
    index.not.empty = which(nchar(body) > 0)
    index.yell = which(nchar(gsub("[A-Z]", "", body[index.not.empty])) < 1)
    return(length(index.yell))
  }
}

isYelling.per = function(msg) {
  body = msg$body
  if(length(body) == 0) return("no body")
  else{
    # remove non-alpha characters
    body = gsub("[^[:alpha:]]", "", body)
    index.not.empty = which(nchar(body) > 0)
    index.yell = which(nchar(gsub("[A-Z]", "", body[index.not.empty])) < 1)
    return(length(index.yell)/length(index.not.empty)) #count empty line or not?
  }
}
#unlist(unname(lapply(emailStruct, isYelling.num)))
#unlist(unname(lapply(emailStruct, isYelling.per)))

```
# Question 3
Check that the hour feature in emailDF gives valid values for all of the email messages. Then perform descriptive analysis to compare this feature for spam and ham, and comment on the possibility of using this feature to classify email.
```{r}
library(RColorBrewer)
cols = brewer.pal(12, "Set3")
#emailDF$hour
isSpamLabs = factor(emailDF$isSpam, labels = c("ham", "spam"))
mosaicplot(table(isSpamLabs, emailDF$hour), main = "",
           xlab = "", ylab = "", color = cols)
barplot(table(isSpamLabs, emailDF$hour),beside = TRUE)

```
# Question 4
Do exercise Q.14 from page 167 of Data Science in R: A Case Studies Approach to Computational Reasoning and Problem Solving, by Deborah Nolan and Duncan Temple Lang.
```{r}
# Original
isRe = function(msg){
  # Can have a Fwd: Re:  ... but we are not looking for this here.
  # We may want to look at In-Reply-To field.
  "Subject" %in% names(msg$header) && 
    length(grep("^[ \t]*Re:", msg$header[["Subject"]])) > 0
}

# Alternative 1
isRe1 = function(msg){
  "Subject" %in% names(msg$header) && 
    (length(grep("^[ \t]*Re:", msg$header[["Subject"]])) > 0 || # why this is start from the beginning, what is ^
       length(grep("^[ \t]*Fwd: Re", msg$header[["Subject"]])) > 0 ) # looking for Fwd: or Fwd:Re: or Re[2] in easyham740
}

# Alternative 2
isRe2 = function(msg){
  "Subject" %in% names(msg$header) && 
       length(grep("Re:", msg$header[["Subject"]])) > 0 
}

idx = which(lapply(emailStruct, isRe)==TRUE)
idx1 =  which(lapply(emailStruct, isRe1)==TRUE)
idx2 = which(lapply(emailStruct, isRe2)==TRUE)
# total
sum(lapply(emailStruct, isRe) == TRUE)
# isSpam
sum(emailDF$isSpam[idx])

sum(lapply(emailStruct, isRe1) == TRUE)
sum(emailDF$isSpam[idx1])
sum(lapply(emailStruct, isRe2) == TRUE)
sum(emailDF$isSpam[idx2])

```